Tree height and hydraulic traits shape growth responses across droughts in a temperate broadleaf forest

### 2019-11-01 manuscript/climate_sensitivity_draft.Rmd
worked on discsussion/ refs

Our analysis indicates that height is the primary factor through which tree size mediates drought response; however, specific identification of the primary physiological mechanism(s) is complicated by the fact that multiple drought-relevant environmental parameters and tree characteristics covary with height. To begin with, taller trees face inherent biophysical challenges in lifting water a greater distance against the effects of gravity and friction (*Ryan et al. 2006-  DOI:10.1111/j.1365-3040.2005.01478.x; McDowell & Allen 2015: DOI: 10.1038/NCLIMATE2641; [McDowell et al. 2011](https://www.springer.com/gp/book/9789400712416)*). They also face different microenvironments, which are partially related to canopy position. Even under non-drought conditions, evaporative demand increases with tree height, which are more closely coupled to the atmosphere  (Fig. 1a-b; **REFS- Jarvis 1984?**; [@bretfeld_plant_2018]). Exposed canopy leaves reach higher temperatures (**REFS**), particularly during drought when solar radiation tends to be higher and less water is available for evaporative cooling of the leaves. Furthermore, daytime CO_2_ concentrations tend to be lower in the canopy than in the understory [@koike_leaf_2001], implying greater water costs of $CO_2$ uptake. Taller trees are more likely to be in dominant canopy positions, although signficant decoupling can be introduced by the configuration of neighboring trees (Fig. 2d) [@muller-landau_testing_2006]. We show that height is a far stronger predictor than crown position (Tables 1,4,5). Our analysis does have the limitation that canopy positions were recored in 2018 and undoubtedly changed for some trees since the 1960s, and we note that $CP$ became an increasingly poor predictor moving from 1999 back to 1966 (Table 4). However, rates of change are relatively slow, and the fact that some trees inevitably did grow towards more dominant positions would bias against the acceptance of *H1.2*. The implication is that dominant crown positions did have a marginally negative influence on $R$, which makes sense in light of the vertical enviromnental gradients described above and agrees with previous studies showing greater drought sensitivity in more exposed trees ([@suarez_factors_2004]; [@scharnweber_confessions_2019]). It is safe to assume that currently suppressed trees have always been suppressed, and their relatively low $R$ (after correcting for height effects) is real, which is consistent with analyses showing that suppressed--and particularly crowded--trees can suffer disproportionately during drought (**REFS**). The observed height-sensitivity of $R$, together with the apparent lack of importance of root water access (*H1.3*), agrees with the concept that physiological limitations to transpiration under drought shift from root water access to the plant-atmosphere intreface as forests age [@bretfeld_plant_2018], such that tall trees--particularly the very tallest--are the most sensitive in mature forests. Additional research comparing drought responses of young and old forest stands, along with and short and tall isolated trees, would be valuable for more clearly disentanging the roles of tree height and crown exposure.

Beyond microenvironment, multiple tree characteristics covary with height. To begin with, multiple tree characteristics that may influence drought resistance scale with $DBH$--or, similarly, $H$--including crown volume, rooting volume, sapwood area, transpiration ... (e.g., [@andersonteixeira_size-related_2015; **REFS--KAT: Sperry scaling model; roots reference; BAAD database?**)]. These work in combination to shape whole-plant hydraulic function (**REFS-KAT: Sperry scaling model, check McDowell scaling paper, Couvreur et al. 2018; others?**), and we cannot expect to statistically identify a single driver that fully encapsulates size-related variation in trees. However, height is arguably the most biophysically meaningful tree-size variable shaping drought response. Beyond whole-plant characteristics, leaf traits vary with height on the tree--sometimes quite dramatically, with upper-canopy leaves ... (**describe differences**) [@couvreur_water_2018; @koike_leaf_2001]. Our analysis focuses on species-level comparisons and does not characterize the role of variation with height. On an individual level, average leaf characteristics of an individual tree would scale with its crown height, with taller individuals having on average more drought-resistant traits--assuming trends for $PLA_{dry}$ and $\pi_{tlp}$, which have not been characterized (***LAWREN, IS THIS TRUE?**), follow the general pattern of increasing drought resistance with height. In not accounting for this, our analysis is most likely biased in a conservative direction when assessing *H2.3* (Table 1)--*i.e.*, accounting traits on an individual level should result in a stronger negative effect of $ln[H]$. Further characterization of leaf hydraulic traits in relation to height and crown exposure would be valuable for enhancing our understanding of the interactive effects of tree height and traits on drought responses [@couvreur_water_2018].

@@

As climate change drives increasing drought in many of the world's forests (*REFS-KAT*), the fate of forests and their climate feedbacks will be shaped by the biophysical and physiological drivers observed here. Large trees have been suffering disrpoportionately in forests around the world (*REFS*), and we here show that this is primarily driven by their height, with some contributions from canopy position. The distinction is important because it suggests that height *per se* makes trees vulnerable, even if their crowns are somewhat protected by neighbors, whereas solitary trees or the dominant trees in young regrowth forests should be less vulnerable. Considering just height and crown position, this would suggest that mature forests would be more vulnerable to drought than young forests with short trees; however, root water access may limit the young forests (Bretfield et al.), and species traits often shift as forests age, with early successional species tending to have lower wood densities and higher hydraulic conductivities that facilitate rapid growth (**REFS in Bretfield-beginning of discussion**). The later successional species at our site (*Fagus grandifolia*, ... ) have a mix of traits conferring drought tolerance and resistance (Table 3), and further research on how hydraulic traits and drought vulnerability change over the course of succession would be valuable for getting at the very signficant question of whether old-growth forests are more vulnerable to drought, and to which types of drought. In the meantime, the results of this study advance our knowledge of the factors conferring drought vulnerability and resistance in a mature forest, opening the door for more accurate forecasting of forest responses to future drought.

>>>>

Our analysis indicates that height is the primary factor through which tree size mediates drought response; however, specific identification of the primary physiological mechanism(s) is complicated by the fact that multiple drought-relevant environmental parameters and tree characteristics covary with height. To begin with, taller trees face inherent biophysical challenges in lifting water a greater distance against the effects of gravity and friction (*Ryan et al. 2006-  DOI:10.1111/j.1365-3040.2005.01478.x; McDowell & Allen 2015: DOI: 10.1038/NCLIMATE2641; [McDowell et al. 2011](https://www.springer.com/gp/book/9789400712416);*)[@couvreur_water_2018. They also face different microenvironments, which are partially related to canopy position. Even under non-drought conditions, evaporative demand increases with tree height, which are more closely coupled to the atmosphere  (Fig. 1a-b; **REFS- Jarvis 1984?**; [@bretfeld_plant_2018]). Exposed canopy leaves reach higher temperatures (**REFS**), particularly during drought when solar radiation tends to be higher and less water is available for evaporative cooling of the leaves. Furthermore, daytime CO_2_ concentrations tend to be lower in the canopy than in the understory [@koike_leaf_2001], implying greater water costs of $CO_2$ uptake. Taller trees are more likely to be in dominant canopy positions, although signficant decoupling can be introduced by the configuration of neighboring trees (Fig. 2d) [@muller-landau_testing_2006]. We show that height is a far stronger predictor than crown position (Tables 1,4,5). Our analysis does have the limitation that canopy positions were recored in 2018 and undoubtedly changed for some trees since the 1960s, and we note that $CP$ became an increasingly poor predictor moving from 1999 back to 1966 (Table 4). However, rates of change are relatively slow, and the fact that some trees inevitably did grow towards more dominant positions would bias against the acceptance of *H1.2*. The implication is that dominant crown positions did have a marginally negative influence on $R$, which makes sense in light of the vertical enviromnental gradients described above and agrees with previous studies showing greater drought sensitivity in more exposed trees ([@suarez_factors_2004]; [@scharnweber_confessions_2019]). It is safe to assume that currently suppressed trees have always been suppressed, and their relatively low $R$ (after correcting for height effects) is real, which is consistent with analyses showing that suppressed--and particularly crowded--trees can suffer disproportionately during drought (**REFS**). The observed height-sensitivity of $R$, together with the apparent lack of importance of root water access (*H1.3*), agrees with the concept that physiological limitations to transpiration under drought shift from root water access to the plant-atmosphere intreface as forests age [@bretfeld_plant_2018], such that tall trees--particularly the very tallest--are the most sensitive in mature forests. Additional research comparing drought responses of young and old forest stands, along with and short and tall isolated trees, would be valuable for more clearly disentanging the roles of tree height and crown exposure.

Beyond microenvironment, multiple tree characteristics covary with height. To begin with, multiple tree characteristics that may influence drought resistance scale with $DBH$--or, similarly, $H$--including leaf area and crown volume, rooting volume, sapwood area, and transpiration (e.g., [@andersonteixeira_size-related_2015;  **REFS--KAT: Sperry scaling model; roots reference; BAAD database?**)]. These work in combination to shape whole-plant hydraulic function (**REFS-KAT: Sperry scaling model, check McDowell scaling paper, Couvreur et al. 2018; others?**)[@couvreur_water_2018], and we cannot expect to statistically identify a single driver that fully encapsulates size-related variation in trees. However, height is arguably the most biophysically meaningful tree-size variable shaping drought response. Beyond whole-plant characteristics, leaf traits vary with height on the tree--sometimes quite dramatically, with upper-canopy leaves tending to be smaller and thicker (higher LMA), with more negative $P50$ and lower hydraulic conductivity [@couvreur_water_2018; @koike_leaf_2001]*[McDowell et al. 2011](https://www.springer.com/gp/book/9789400712416)*. Our analysis focuses on species-level comparisons and does not characterize the role of variation with height. On an individual level, average leaf characteristics of an individual tree would scale with its crown height, with taller individuals having on average more drought-resistant traits--assuming trends for $PLA_{dry}$ and $\pi_{tlp}$, which have not been characterized (***LAWREN, IS THIS TRUE?**), follow the general pattern of increasing drought resistance with height. In not accounting for this, our analysis is most likely biased in a conservative direction when assessing *H2.3* (Table 1)--*i.e.*, accounting traits on an individual level should result in a stronger negative effect of $ln[H]$. Further characterization of leaf hydraulic traits in relation to height and crown exposure would be valuable for enhancing our understanding of the interactive effects of tree height and traits on drought responses [@couvreur_water_2018].

@@

As climate change drives increasing drought in many of the world's forests (*REFS-KAT*), the fate of forests and their climate feedbacks will be shaped by the biophysical and physiological drivers observed here. Large trees have been suffering disrpoportionately in forests around the world [@bennett_larger_2015; @stovall_tree_2019], and we here show that this is primarily driven by their height, with some contributions from canopy position. The distinction is important because it suggests that height *per se* makes trees vulnerable, even if their crowns are somewhat protected by neighbors, whereas solitary trees or the dominant trees in young regrowth forests should be less vulnerable. Considering just height and crown position, this would suggest that mature forests would be more vulnerable to drought than young forests with short trees; however, root water access may limit the young forests (Bretfield et al.), and species traits often shift as forests age, with early successional species tending to have lower wood densities and higher hydraulic conductivities that facilitate rapid growth (**REFS in Bretfield-beginning of discussion**). The later successional species at our site (*Fagus grandifolia*, ... ) have a mix of traits conferring drought tolerance and resistance (Table 3), and further research on how hydraulic traits and drought vulnerability change over the course of succession would be valuable for getting at the very signficant question of whether old-growth forests are more vulnerable to drought, and to which types of drought. In the meantime, the results of this study advance our knowledge of the factors conferring drought vulnerability and resistance in a mature forest, opening the door for more accurate forecasting of forest responses to future drought.
